---
title: "Reproducible Research Project 1"
output: html_document
---
1. LOADING AND PREPROCESSING DATA

The required R packages are read into R:

```{r}
library(plyr)
library(dplyr)
library(ggplot2)
```

The data can be found in a zipped file at: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip
The data should be extracted and saved in the R working directory as activity.csv.

The raw data are then read into R and saved as "data"

```{r}
data <- read.csv("activity.csv")
```

The data set is then process to a useful format.

```{r}
data<-mutate(data,day=weekdays(as.Date(date,'%Y-%m-%d')))
data<-mutate(data,v1=as.numeric(steps))

mydate<-data$date
mydate<-as.Date(mydate, format = "%Y-%m-%d")
data<-cbind(data,mydate)

data<-select(data,v1,mydate,interval,day)
data<-rename(data,steps=v1)
data<-rename(data,date=mydate)
```

2. TOTAL, MEAN AND MEDIAN STEPS FOR EACH DAY

The daily totals,and averages and median for each day's 5 minute interval are calculated:

```{r}
day.tot <- ddply(data, c("date"), summarise,
                      sum  = sum(steps))

day.mean <- ddply(data, c("date"), summarise,
                 mean  = mean(steps))

day.median <-ddply(data, c("date"), summarise,
                   median  = median(steps))

tot.mean.median<-cbind(day.tot,day.mean$mean,day.median$median)

tot.mean.median
```

A histogram of the total daily steps is plotted:

```{r, echo=TRUE}
hist(day.tot$sum,xlab="Total Steps per Day",main="Histogram of Daily Steps")
```

3. AVERAGE DAILY ACTIVITY PATTERM

A time series plot including NA (breaks in the line) data is shown below.

```{r, echo=TRUE}
plot(day.mean$date, day.mean$mean,type="l",xlab="Date",ylab="Average Daily Steps")
```

The 5 minute interval with the highest number of steps on averages is:

```{r}
comp.data<-data[complete.cases(data),]
uncomp.data<-data[!complete.cases(data),]

int.mean <- ddply(comp.data, c("interval"), summarise,
                  mean  = mean(steps))

arrange(int.mean,desc(mean))[1,]
```

4. IMPUTING MISSING VALUES
Missing data (NAs) are replaced with the average values for the corresponding 5 minute interval.

```{r}
uncomp.data<-data[!complete.cases(data),]
b<-merge(int.mean,uncomp.data,by="interval")
b<-select(b,interval,mean,date,day)
b<-rename(b,steps=mean)
impute.data<-rbind(b,comp.data)
```

A histogram of daily step totals with the imputed data is given below:

```{r, echo=TRUE}
day.imptot <- ddply(impute.data, c("date"), summarise,
                 sum  = sum(steps))

hist(day.imptot$sum,main="Histogram of Steps After Imputing Missing Data",xlab="Total Daily Steps")
```

Using the average steps for each interval leads to the following means and medians for each day.

```{r}
day.impmean <- ddply(impute.data, c("date"), summarise,
                    mean  = mean(steps))

day.impmedian <- ddply(impute.data, c("date"), summarise,
                    median  = median(steps))
com.imp<-cbind(day.impmean,day.impmedian$median)
com.imp
```

The median values given for days with many missing data points are inflated. A better imputing stategy should be considered.
Totals for and means are slightly changed by imputing values for missing data.

5. WEEKDAY AND WEEKEND VALUES

```{r,echo=TRUE}
wend<-subset(impute.data,day=="Sunday"|day=="Saturday")
wday<-subset(impute.data,day=="Monday"|day=="Tuesday"|day=="Wednesday"|day=="Thursday"|day=="Friday")

we<-ddply(wend,c("interval"),summarise,
          mean = mean(steps))
wd<-ddply(wday,c("interval"),summarise,
          mean = mean(steps))

weekend<-mutate(we,wk="Weekend")
weekday<-mutate(wd,wk="Weekday")

wkdata<-rbind(weekend,weekday)

qplot(interval,mean,data=wkdata,geom=c("line"),facets=.~wk)
```

